A Short Demo of data.table
* Base R
* The Data.Table Package
* Creating a data.table
* Creating a data.table 2
* data.table Commands
* The i= Argument
* The i= Argument
* Data.Table: the “j” argument
* Data.Table: the “j” argument
* Using Integers and Characters in the j argument
* More Complex j arguments
* The j expression
* The j argument, complex expressions
* The j expression, reducing results
* Using the j argument for variable transformations
* Evaluating expressions by groups
* Evaluating expressions by groups
* Evaluating expressions by groups
* Sorting a Data.table
* Regression by Group
* Conclusion

Base R
In base R, we use the ‘dataframe’ data structure to organize data.

We also use read.csv() to import data. This can be slow though…

You may also be familiar with the tidyverse, or specifically dplyr for managing data in R.

Possible causes for base R being slow:

** Bad r code (not pre-allocating result vectors)
** Lazy copying when manipulating dataframes (like renaming)
** Slower sort algorithms
** Not leveraging sort order

The Data.Table Package
* A different approach to dataframes in R
* data.table replaces the data.frame
* A data.table object is also a data.frame
** You can use a data.table anywhere you can use a data.frame
* Goals of the data.table package
** Reduce time to write code through sensical syntax (opinion)
** Reduce the time code takes to run through efficient copying ordered grouping and ordered joins (fact)
** Some of the computing time gains depend on sorting (setting keys) the data.table then leveraging the sort order for joins, group by calculations, and subsetting

Creating a data.table
* We can directly create a data.table similar to a data.frame
* We can easily convert a data.frame to a data.table
* We can use fread() to read in structured text as a data.table, this replacesread.csv()

```{r}
library(data.table) 
 
## We can directly create a data.table just like a dataframe 
dt1 <-  
  data.table(x1=1:10, 
             x2=10:1) 
dt1 
##     x1 x2
##  1:  1 10
##  2:  2  9
##  3:  3  8
##  4:  4  7
##  5:  5  6
##  6:  6  5
##  7:  7  4
##  8:  8  3
##  9:  9  2
## 10: 10  1
```
```{r}
str(dt1) 
## Classes 'data.table' and 'data.frame':   10 obs. of  2 variables:
##  $ x1: int  1 2 3 4 5 6 7 8 9 10
##  $ x2: int  10 9 8 7 6 5 4 3 2 1
##  - attr(*, ".internal.selfref")=<externalptr>
```
```{r}
is.data.frame(dt1) 
## [1] TRUE

```

```{r}
is.data.table(dt1) 
## [1] TRUE

```

Creating a data.table 2
* We can easily convert a data.frame to a data.table
* We can use fread to read in structured text as a data.table (this replaces read.csv)

```{r}

library(data.table) 
library(plyr) # for the baseball data
 
## We can convert existing dataframe to dt 
data(baseball) 
str(baseball) 
## 'data.frame':    21699 obs. of  22 variables:
##  $ id   : chr  "ansonca01" "forceda01" "mathebo01" "startjo01" ...
##  $ year : int  1871 1871 1871 1871 1871 1871 1871 1872 1872 1872 ...
##  $ stint: int  1 1 1 1 1 1 1 1 1 1 ...
##  $ team : chr  "RC1" "WS3" "FW1" "NY2" ...
##  $ lg   : chr  "" "" "" "" ...
##  $ g    : int  25 32 19 33 29 29 29 46 37 25 ...
##  $ ab   : int  120 162 89 161 128 146 145 217 174 130 ...
##  $ r    : int  29 45 15 35 35 40 36 60 26 40 ...
##  $ h    : int  39 45 24 58 45 47 37 90 46 53 ...
##  $ X2b  : int  11 9 3 5 3 6 5 10 3 11 ...
##  $ X3b  : int  3 4 1 1 7 5 7 7 0 0 ...
##  $ hr   : int  0 0 0 1 3 1 2 0 0 0 ...
##  $ rbi  : int  16 29 10 34 23 21 23 50 15 16 ...
##  $ sb   : int  6 8 2 4 3 2 2 6 0 2 ...
##  $ cs   : int  2 0 1 2 1 2 2 6 1 2 ...
##  $ bb   : int  2 4 2 3 1 4 9 16 1 1 ...
##  $ so   : int  1 0 0 0 0 1 1 3 1 0 ...
##  $ ibb  : int  NA NA NA NA NA NA NA NA NA NA ...
##  $ hbp  : int  NA NA NA NA NA NA NA NA NA NA ...
##  $ sh   : int  NA NA NA NA NA NA NA NA NA NA ...
##  $ sf   : int  NA NA NA NA NA NA NA NA NA NA ...
##  $ gidp : int  NA NA NA NA NA NA NA NA NA NA ...

```

```{r}

baseball_dt <-  
  data.table(baseball) 
 
## Fast File Read 
cohort <- fread("patient_demo.csv") 
```

```{r}
## how much faster?
system.time(cohort <- read.csv("patient_demo.csv"))
##    user  system elapsed 
##   0.177   0.004   0.182

```

```{r}

library(readr)
system.time(cohort <- read_csv("patient_demo.csv"))
## 
## ── Column specification ────────────────────────────────────────────────────────
## cols(
##   patient_id = col_double(),
##   hosp_id = col_double(),
##   age = col_double(),
##   gender = col_character(),
##   mortality = col_character(),
##   SES = col_character()
## )
##    user  system elapsed 
##   0.212   0.012   0.328
system.time(cohort <- fread("patient_demo.csv"))
##    user  system elapsed 
##   0.055   0.000   0.031

```


data.table Commands
* Recall a data.table is also a data.frame so all the data.frame methods work
* Data.table prints only the first 5 and last 5 rows of a table if the table is larger than 100 rows

```{r}

## list all data.tables and their size 
tables() 
##           NAME    NROW NCOL MB                                        COLS KEY
## 1: baseball_dt  21,699   22  2                 id,year,stint,team,lg,g,...    
## 2:      cohort 100,000    6  3 patient_id,hosp_id,age,gender,mortality,SES    
## 3:         dt1      10    2  0                                       x1,x2    
## Total: 5MB
## Confirm this is a data.table (and a data frame) 
```

```{r}
class(cohort) 
## [1] "data.table" "data.frame"
## the usual methods for data.frames all work for  
```

```{r}
summary(cohort) 
##    patient_id        hosp_id            age            gender         
##  Min.   :100000   Min.   :  1.00   Min.   : 22.00   Length:100000     
##  1st Qu.:125000   1st Qu.: 25.00   1st Qu.: 58.00   Class :character  
##  Median :150000   Median : 51.00   Median : 65.00   Mode  :character  
##  Mean   :150000   Mean   : 50.55   Mean   : 64.93                     
##  3rd Qu.:174999   3rd Qu.: 75.00   3rd Qu.: 72.00                     
##  Max.   :199999   Max.   :100.00   Max.   :109.00                     
##   mortality             SES           
##  Length:100000      Length:100000     
##  Class :character   Class :character  
##  Mode  :character   Mode  :character  
##                                       
##                                       
## 
```

```{r}
## Data.tables print the first and last 5 rows 
cohort 
##         patient_id hosp_id age gender mortality  SES
##      1:     100000      77  70   Male       Yes  Med
##      2:     100001      45  64   Male        No  Low
##      3:     100002       5  76   Male        No  Med
##      4:     100003      13  51   Male        No High
##      5:     100004      70  76   Male        No  Low
##     ---                                             
##  99996:     199995      60  66   Male        No  Med
##  99997:     199996      14  71   Male        No  Low
##  99998:     199997      39  73   Male       Yes  Med
##  99999:     199998      21  60 Female        No High
## 100000:     199999      81  61   Male        No  Low
```

The i= Argument
* We will use the square brackets to interact with data.tables (the subset operator)
* Forget everything you know about the square brackets (sorry!), it is different in data.table
* The first argument to the square brackets is i=, which we use to subset the rows of the data.table
* This can be an integer vector or a Boolean vector
* Any references to variables will be first resolved within the data.table environment
```

```{r}
cohort[] # prints all 
##         patient_id hosp_id age gender mortality  SES
##      1:     100000      77  70   Male       Yes  Med
##      2:     100001      45  64   Male        No  Low
##      3:     100002       5  76   Male        No  Med
##      4:     100003      13  51   Male        No High
##      5:     100004      70  76   Male        No  Low
##     ---                                             
##  99996:     199995      60  66   Male        No  Med
##  99997:     199996      14  71   Male        No  Low
##  99998:     199997      39  73   Male       Yes  Med
##  99999:     199998      21  60 Female        No High
## 100000:     199999      81  61   Male        No  Low
```

```{r}
## the first row 
cohort[i=1] 
##    patient_id hosp_id age gender mortality SES
## 1:     100000      77  70   Male       Yes Med
```

```{r}
## we usually don't bother naming the argument with i= 
## Rows 10 to 20 
cohort[10:20] 
##     patient_id hosp_id age gender mortality  SES
##  1:     100009      64  71   Male        No  Low
##  2:     100010      43  83   Male        No  Low
##  3:     100011      62  66   Male        No  Low
##  4:     100012      54  58   Male        No  Med
##  5:     100013      69  82   Male        No  Low
##  6:     100014      20  72   Male        No  Med
##  7:     100015      46  48   Male        No High
##  8:     100016      19  71   Male        No High
##  9:     100017       1  70   Male        No  Med
## 10:     100018      81  60   Male        No  Med
## 11:     100019       5  76   Male        No  Med
```
The i= Argument

```{r}
## all patients older than 50 
cohort[age>50] 
##        patient_id hosp_id age gender mortality  SES
##     1:     100000      77  70   Male       Yes  Med
##     2:     100001      45  64   Male        No  Low
##     3:     100002       5  76   Male        No  Med
##     4:     100003      13  51   Male        No High
##     5:     100004      70  76   Male        No  Low
##    ---                                             
## 92508:     199995      60  66   Male        No  Med
## 92509:     199996      14  71   Male        No  Low
## 92510:     199997      39  73   Male       Yes  Med
## 92511:     199998      21  60 Female        No High
## 92512:     199999      81  61   Male        No  Low
```

```{r}
## all patients older than 50 and male 
cohort[age>50 & gender == "Male"] 
##        patient_id hosp_id age gender mortality  SES
##     1:     100000      77  70   Male       Yes  Med
##     2:     100001      45  64   Male        No  Low
##     3:     100002       5  76   Male        No  Med
##     4:     100003      13  51   Male        No High
##     5:     100004      70  76   Male        No  Low
##    ---                                             
## 83291:     199994      65  85   Male        No  Med
## 83292:     199995      60  66   Male        No  Med
## 83293:     199996      14  71   Male        No  Low
## 83294:     199997      39  73   Male       Yes  Med
## 83295:     199999      81  61   Male        No  Low
```

Question: Select the first 10 rows of the baseball data

Answer

```{r}
baseball_dt[1:10] 
##            id year stint team lg  g  ab  r  h X2b X3b hr rbi sb cs bb so ibb
##  1: ansonca01 1871     1  RC1    25 120 29 39  11   3  0  16  6  2  2  1  NA
##  2: forceda01 1871     1  WS3    32 162 45 45   9   4  0  29  8  0  4  0  NA
##  3: mathebo01 1871     1  FW1    19  89 15 24   3   1  0  10  2  1  2  0  NA
##  4: startjo01 1871     1  NY2    33 161 35 58   5   1  1  34  4  2  3  0  NA
##  5: suttoez01 1871     1  CL1    29 128 35 45   3   7  3  23  3  1  1  0  NA
##  6: whitede01 1871     1  CL1    29 146 40 47   6   5  1  21  2  2  4  1  NA
##  7:  yorkto01 1871     1  TRO    29 145 36 37   5   7  2  23  2  2  9  1  NA
##  8: ansonca01 1872     1  PH1    46 217 60 90  10   7  0  50  6  6 16  3  NA
##  9: burdoja01 1872     1  BR2    37 174 26 46   3   0  0  15  0  1  1  1  NA
## 10: forceda01 1872     1  TRO    25 130 40 53  11   0  0  16  2  2  1  0  NA
##     hbp sh sf gidp
##  1:  NA NA NA   NA
##  2:  NA NA NA   NA
##  3:  NA NA NA   NA
##  4:  NA NA NA   NA
##  5:  NA NA NA   NA
##  6:  NA NA NA   NA
##  7:  NA NA NA   NA
##  8:  NA NA NA   NA
##  9:  NA NA NA   NA
## 10:  NA NA NA   NA
```

Question: Select the rows after the year 2000 for the baseball data

Answer
```{r}
baseball_dt[year > 2000] 
##              id year stint team lg   g  ab   r   h X2b X3b hr rbi sb cs bb so
##    1: alomaro01 2001     1  CLE AL 157 575 113 193  34  12 20 100 30  6 80 71
##    2: alomasa02 2001     1  CHA AL  70 220  17  54   8   1  4  21  1  2 12 17
##    3: anderbr01 2001     1  BAL AL 131 430  50  87  12   3  8  45 12  4 60 77
##    4: baineha01 2001     1  CHA AL  32  84   3  11   1   0  0   6  0  0  8 16
##    5:  bellda01 2001     1  SEA AL 135 470  62 122  28   0 15  64  2  1 28 59
##   ---                                                                        
## 1266: benitar01 2007     2  FLO NL  34   0   0   0   0   0  0   0  0  0  0  0
## 1267: benitar01 2007     1  SFN NL  19   0   0   0   0   0  0   0  0  0  0  0
## 1268: ausmubr01 2007     1  HOU NL 117 349  38  82  16   3  3  25  6  1 37 74
## 1269:  aloumo01 2007     1  NYN NL  87 328  51 112  19   1 13  49  3  0 27 30
## 1270: alomasa02 2007     1  NYN NL   8  22   1   3   1   0  0   0  0  0  0  3
##       ibb hbp sh sf gidp
##    1:   5   4  9  9    9
##    2:   1   2  3  2    6
##    3:   4   8  2  1    3
##    4:   0   0  0  2    2
##    5:   1   3  5  4    8
##   ---                   
## 1266:   0   0  0  0    0
## 1267:   0   0  0  0    0
## 1268:   3   6  4  1   11
## 1269:   5   2  0  3   13
## 1270:   0   0  0  0    0
```
Question: Select the rows with team equal to CLE or NYN
Answer

```{r}
baseball_dt[team %in% c("NYN","CLE")] 
##              id year stint team lg   g  ab  r   h X2b X3b hr rbi sb cs bb  so
##    1: hallmbi01 1901     1  CLE AL   5  19  2   4   0   0  0   3  0 NA  2  NA
##    2: mooreea01 1901     1  CLE AL  31  99  5  16   0   0  0   6  1 NA  6  NA
##    3: weyhigu01 1901     1  CLE AL   2   5  0   0   0   0  0   0  0 NA  0  NA
##    4: hickmch01 1902     2  CLE AL 102 426 61 161  31  11  8  94  8 NA 12  NA
##    5: lajoina01 1902     2  CLE AL  86 348 81 132  35   5  7  64 19 NA 19  NA
##   ---                                                                        
## 1461: easleda01 2007     1  NYN NL  76 193 24  54   6   0 10  26  0  1 19  35
## 1462: delgaca01 2007     1  NYN NL 139 538 71 139  30   0 24  87  4  0 52 118
## 1463: coninje01 2007     2  NYN NL  21  41  2   8   2   0  0   5  0  0  7   8
## 1464:  aloumo01 2007     1  NYN NL  87 328 51 112  19   1 13  49  3  0 27  30
## 1465: alomasa02 2007     1  NYN NL   8  22  1   3   1   0  0   0  0  0  0   3
##       ibb hbp sh sf gidp
##    1:  NA   0  0 NA   NA
##    2:  NA   0  3 NA   NA
##    3:  NA   0  0 NA   NA
##    4:  NA   3  8 NA   NA
##    5:  NA   6  8 NA   NA
##   ---                   
## 1461:   1   5  0  1    2
## 1462:   8  11  0  6   12
## 1463:   2   0  1  1    1
## 1464:   5   2  0  3   13
## 1465:   0   0  0  0    0
```


Data.Table: the “j” argument
* The second argument to the square brackets is the j argument
* We leave the first argument blank (the i argument) if we want all the rows and just put a comma
* The j argument can be any valid R expression
* If the j argument resolves to a list, a data.table is returned

```{r}
## Returns age as a vector 
cohort[,j=age][1:10] 
##  [1] 70 64 76 51 76 60 55 66 75 71

```

```{r}
## we typically omit the j= piece 
cohort[,age][1:10]
##  [1] 70 64 76 51 76 60 55 66 75 71
```

```{r}
## Adding list() returns a data.table 
cohort[,list(age)] 
##         age
##      1:  70
##      2:  64
##      3:  76
##      4:  51
##      5:  76
##     ---    
##  99996:  66
##  99997:  71
##  99998:  73
##  99999:  60
## 100000:  61

```

```{r}
cohort[,list(age,patient_id)] 
##         age patient_id
##      1:  70     100000
##      2:  64     100001
##      3:  76     100002
##      4:  51     100003
##      5:  76     100004
##     ---               
##  99996:  66     199995
##  99997:  71     199996
##  99998:  73     199997
##  99999:  60     199998
## 100000:  61     199999
## use . as a shortcut for list here 
```

```{r}
cohort[,.(age,patient_id)] 
##         age patient_id
##      1:  70     100000
##      2:  64     100001
##      3:  76     100002
##      4:  51     100003
##      5:  76     100004
##     ---               
##  99996:  66     199995
##  99997:  71     199996
##  99998:  73     199997
##  99999:  60     199998
## 100000:  61     199999
```

Data.Table: the “j” argument
Question: From the Baseball data, return a data.table with the rows from after the year 2000 with only the id, year and team columns using the list() command

Answer
```{r}
baseball_dt[year > 2000,list(id,year,team)] 
##              id year team
##    1: alomaro01 2001  CLE
##    2: alomasa02 2001  CHA
##    3: anderbr01 2001  BAL
##    4: baineha01 2001  CHA
##    5:  bellda01 2001  SEA
##   ---                    
## 1266: benitar01 2007  FLO
## 1267: benitar01 2007  SFN
## 1268: ausmubr01 2007  HOU
## 1269:  aloumo01 2007  NYN
## 1270: alomasa02 2007  NYN
```


Question: Repeat this using the . to replace the list command
Answer

```{r}
baseball_dt[year > 2000,.(id,year,team)] 
##              id year team
##    1: alomaro01 2001  CLE
##    2: alomasa02 2001  CHA
##    3: anderbr01 2001  BAL
##    4: baineha01 2001  CHA
##    5:  bellda01 2001  SEA
##   ---                    
## 1266: benitar01 2007  FLO
## 1267: benitar01 2007  SFN
## 1268: ausmubr01 2007  HOU
## 1269:  aloumo01 2007  NYN
## 1270: alomasa02 2007  NYN
```


Question: How many rows meet this condition?
Answer
```{r}
## How many rows? 
nrow(baseball_dt[year > 2000,.(id,year,team)]) 
## [1] 1270
```

Using Integers and Characters in the j argument
* In base R we can use integers or characters to reference columns as the second argument following [
* The same behavior is true in data.table (this was not always the case in prior versions of data.table)
* However, one difference is the class of the return object. Notice data.table returns a data.table, while the data.frame would return a vector instead of a dataframe.

```{r}
# These work as expected 
cohort[,1] 
##         patient_id
##      1:     100000
##      2:     100001
##      3:     100002
##      4:     100003
##      5:     100004
##     ---           
##  99996:     199995
##  99997:     199996
##  99998:     199997
##  99999:     199998
## 100000:     199999
```

```{r}
cohort[,"age"] 
##         age
##      1:  70
##      2:  64
##      3:  76
##      4:  51
##      5:  76
##     ---    
##  99996:  66
##  99997:  71
##  99998:  73
##  99999:  60
## 100000:  61
```
```{r}
# These do not!
var_num <- 1
cohort[,var_num] 
## Error in `[.data.table`(cohort, , var_num): j (the 2nd argument inside [...]) is a single symbol but column name 'var_num' is not found. Perhaps you intended DT[, ..var_num]. This difference to data.frame is deliberate and explained in FAQ 1.1.


```{r}
cohort[,..var_num] # works
##         patient_id
##      1:     100000
##      2:     100001
##      3:     100002
##      4:     100003
##      5:     100004
##     ---           
##  99996:     199995
##  99997:     199996
##  99998:     199997
##  99999:     199998
## 100000:     199999
```

```{r}
var_name <- "age"
cohort[,var_name] 
## Error in `[.data.table`(cohort, , var_name): j (the 2nd argument inside [...]) is a single symbol but column name 'var_name' is not found. Perhaps you intended DT[, ..var_name]. This difference to data.frame is deliberate
```

```{r}
cohort[,..var_name] # works
##         age
##      1:  70
##      2:  64
##      3:  76
##      4:  51
##      5:  76
##     ---    
##  99996:  66
##  99997:  71
##  99998:  73
##  99999:  60
## 100000:  61
```

More Complex j arguments
* As stated the j argument can be any valid R expression
* If we return a list of items, it will be coerced into a data.table
* We can name the items of the list, and it will be inherited to column names

```{r}
## here is the average and SD of age 
cohort[,mean(age)] 
## [1] 64.93313
```

```{r}
cohort[,sd(age)] 
## [1] 10.00578
## Returned as a data.table 
```

```{r}
cohort[,list(mean(age), 
             sd(age))] 
##          V1       V2
## 1: 64.93313 10.00578
```

```{r}
## Name each column in the result
cohort[,list(avg=mean(age), 
             stdev=sd(age))] 
##         avg    stdev
## 1: 64.93313 10.00578
```


The j expression
Question: Calculate the average home runs (hr variable) and return the result as a data.table 

```{r}
Answer
baseball_dt[,list(mean(hr))] 
##          V1
## 1: 5.234204
```

Question: Calculate the total number of at bats (ab variable) and the total number of rbi’s (rbi variable) and the total number of players with a team value of “HOU”. Name these columns in the resulting data.table
Answer
```{r}
baseball_dt[,list(tot_ab = sum(ab), 
                  tot_rbi = sum(rbi,na.rm=T), 
                  tot_hous = sum(team=="HOU"))] 
##     tot_ab tot_rbi tot_hous
## 1: 4891061  641644      497
```

The j argument, complex expressions
We can pass in more complex expressions in the j expression
As before, if the result is not a simple list, it will not be a data.table
This should remind you of thewith() command in base r dataframes

```{r}
## here is a table of gender by mortality 
cohort[,table(gender,mortality)] 
##         mortality
## gender      No   Yes
##   Female  8399  1533
##   Male   73612 16456
## With row proportions 
```

```{r}
cohort[,prop.table(table(gender,mortality),1)] 
##         mortality
## gender          No       Yes
##   Female 0.8456504 0.1543496
##   Male   0.8172936 0.1827064
```

```{r}
## chi-square test on gender and mortality 
cohort[,chisq.test(gender,mortality)] 
## 
##  Pearson's Chi-squared test with Yates' continuity correction
## 
## data:  gender and mortality
## X-squared = 48.566, df = 1, p-value = 3.194e-12
```

The j expression, reducing results
As an example of a more complex argument to j lets perform a logistic regression on this dataset

```{r}
## logistic regression on mortality 
mod_result_1 <-  
  cohort[,glm(mortality == "Yes" ~ age + gender + SES, 
              family = "binomial")] 
summary(mod_result_1) 
## 
## Call:
## glm(formula = mortality == "Yes" ~ age + gender + SES, family = "binomial")
## 
## Deviance Residuals: 
##     Min       1Q   Median       3Q      Max  
## -0.9126  -0.6536  -0.6099  -0.5495   2.1987  
## 
## Coefficients:
##               Estimate Std. Error z value Pr(>|z|)    
## (Intercept) -2.9637216  0.0640235 -46.291  < 2e-16 ***
## age          0.0189530  0.0008296  22.846  < 2e-16 ***
## genderMale   0.2065857  0.0291515   7.087 1.37e-12 ***
## SESLow       0.0305045  0.0232105   1.314    0.189    
## SESMed       0.0144423  0.0223347   0.647    0.518    
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## (Dispersion parameter for binomial family taken to be 1)
## 
##     Null deviance: 94245  on 99999  degrees of freedom
## Residual deviance: 93667  on 99995  degrees of freedom
## AIC: 93677
## 
## Number of Fisher Scoring iterations: 4
```

```{r}
## all in one step 
cohort[,summary(glm(mortality == "Yes" ~ age + gender + SES, 
                    family = "binomial"))] 
## 
## Call:
## glm(formula = mortality == "Yes" ~ age + gender + SES, family = "binomial")
## 
## Deviance Residuals: 
##     Min       1Q   Median       3Q      Max  
## -0.9126  -0.6536  -0.6099  -0.5495   2.1987  
## 
## Coefficients:
##               Estimate Std. Error z value Pr(>|z|)    
## (Intercept) -2.9637216  0.0640235 -46.291  < 2e-16 ***
## age          0.0189530  0.0008296  22.846  < 2e-16 ***
## genderMale   0.2065857  0.0291515   7.087 1.37e-12 ***
## SESLow       0.0305045  0.0232105   1.314    0.189    
## SESMed       0.0144423  0.0223347   0.647    0.518    
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## (Dispersion parameter for binomial family taken to be 1)
## 
##     Null deviance: 94245  on 99999  degrees of freedom
## Residual deviance: 93667  on 99995  degrees of freedom
## AIC: 93677
## 
## Number of Fisher Scoring iterations: 4
```

Using the j argument for variable transformations
We may wish to have a transformed version of a variable
We can name simply make a list of the variables we want, and add in the transformations

```{r}
## Lets select patient_id, age, and add log age 
cohort[,list(patient_id, 
             age, 
             l_age=log(age))] 
##         patient_id age    l_age
##      1:     100000  70 4.248495
##      2:     100001  64 4.158883
##      3:     100002  76 4.330733
##      4:     100003  51 3.931826
##      5:     100004  76 4.330733
##     ---                        
##  99996:     199995  66 4.189655
##  99997:     199996  71 4.262680
##  99998:     199997  73 4.290459
##  99999:     199998  60 4.094345
## 100000:     199999  61 4.110874
```

Evaluating expressions by groups
If we specify the by= argument, the j argument will be evaluated within each group specified
This is an especially fast way to evaluate expressions by group levels
The by= variable is automatically returned as part of the expression

```{r}
## Average age by different gender levels 
cohort[,mean(age),by=gender] 
##    gender       V1
## 1:   Male 64.91831
## 2: Female 65.06756
```

```{r}
## Use a named list to name the variable 
cohort[,list(avg_age=mean(age)), 
       by=gender] 
##    gender  avg_age
## 1:   Male 64.91831
## 2: Female 65.06756
```

```{r}
## Return multiple columns 
cohort[,list(avg_age=mean(age), 
             sd_age=sd(age)), 
       by=gender] 
##    gender  avg_age    sd_age
## 1:   Male 64.91831 10.007413
## 2: Female 65.06756  9.990478
```

Evaluating expressions by groups
Question: For the rows after the year 2000, calculate the average and total rbi’s for each team

Answer
```{r}
baseball_dt[year > 2000, 
            list(avg_rbi=mean(rbi,na.rm=T), 
                 tot_rbi=sum(rbi,na.rm=T)), 
            by=team] 
##     team   avg_rbi tot_rbi
##  1:  CLE 31.435897    1226
##  2:  CHA 35.216216    1303
##  3:  BAL 30.270270    1120
##  4:  SEA 26.783333    1607
##  5:  BOS 18.507463    1240
##  6:  TEX 25.697674    1105
##  7:  TBA 27.555556     496
##  8:  NYA 20.189873    1595
##  9:  TOR 44.785714    1254
## 10:  DET 31.150000     623
## 11:  OAK 19.470588     331
## 12:  KCA 17.741935     550
## 13:  MIN  8.730769     227
## 14:  ANA  1.727273      19
## 15:  HOU 34.094340    1807
## 16:  NYN 18.658537    1530
## 17:  LAN 25.589286    1433
## 18:  COL 25.222222    1135
## 19:  ARI 26.507692    1723
## 20:  SLN 27.711111    1247
## 21:  SFN 35.312500    2260
## 22:  ATL 22.104167    1061
## 23:  MIL 27.384615     712
## 24:  PHI 14.318182     630
## 25:  MON 22.777778     205
## 26:  CHN 31.851852    1720
## 27:  PIT 16.818182     370
## 28:  FLO 21.193548     657
## 29:  SDN 19.044776    1276
## 30:  CIN 22.000000     770
## 31:  WAS 13.222222     119
## 32:  LAA 27.000000      54
```

Question: For only the rows after the year 2000, calculate the average rbi’s for each league variable (lg)
Answer
```{r}
##     team   avg_rbi tot_rbi
baseball_dt[year > 2000, 
            list(avg_rbi=mean(rbi,na.rm=T), 
                 tot_rbi=sum(rbi,na.rm=T)), 
            by=lg] 
##    lg  avg_rbi tot_rbi
## 1: AL 24.75728   12750
## 2: NL 24.70861   18655
```


Evaluating expressions by groups
We may want to group by more than one level
In that case, we can pass in either a list of column names as objects, or a character vector of column names
If we pass in a list, we can create a new expression level in the list

```{r}
## multiple by variables using list 
cohort[,list(age=mean(age)),
       by=list(gender,SES)] 
##    gender  SES      age
## 1:   Male  Med 64.83112
## 2:   Male  Low 64.95914
## 3:   Male High 65.04295
## 4: Female  Med 64.92555
## 5: Female High 65.23148
## 6: Female  Low 65.16029
```

```{r}
## multiple by variables using character vector 
cohort[,list(age=mean(age)),
       by=c("gender","SES")] 
##    gender  SES      age
## 1:   Male  Med 64.83112
## 2:   Male  Low 64.95914
## 3:   Male High 65.04295
## 4: Female  Med 64.92555
## 5: Female High 65.23148
## 6: Female  Low 65.16029
```

```{r}
## Use an expression for a by level 
cohort[,list(age=mean(age)), 
       by=list(gender,young=age<50)] 
##    gender young      age
## 1:   Male FALSE 66.22100
## 2: Female FALSE 66.33298
## 3:   Male  TRUE 45.22405
## 4: Female  TRUE 44.99491
```

Sorting a Data.table
* We often wish to sort the results of an expression such as the one we just saw
* We can use setorder() command to sort a data.table in place
* Rather than set the order in place, we can also return the data.table in a new order by calling the order() command, then passing it into the i= argument.

```{r}
res1 <-  
  baseball_dt[year > 2000, 
            list(avg_rbi=mean(rbi,na.rm=T), 
                 tot_rbi=sum(rbi,na.rm=T)), 
            by=team] 
setorder(res1,avg_rbi) 
res1
##     team   avg_rbi tot_rbi
##  1:  ANA  1.727273      19
##  2:  MIN  8.730769     227
##  3:  WAS 13.222222     119
##  4:  PHI 14.318182     630
##  5:  PIT 16.818182     370
##  6:  KCA 17.741935     550
##  7:  BOS 18.507463    1240
##  8:  NYN 18.658537    1530
##  9:  SDN 19.044776    1276
## 10:  OAK 19.470588     331
## 11:  NYA 20.189873    1595
## 12:  FLO 21.193548     657
## 13:  CIN 22.000000     770
## 14:  ATL 22.104167    1061
## 15:  MON 22.777778     205
## 16:  COL 25.222222    1135
## 17:  LAN 25.589286    1433
## 18:  TEX 25.697674    1105
## 19:  ARI 26.507692    1723
## 20:  SEA 26.783333    1607
## 21:  LAA 27.000000      54
## 22:  MIL 27.384615     712
## 23:  TBA 27.555556     496
## 24:  SLN 27.711111    1247
## 25:  BAL 30.270270    1120
## 26:  DET 31.150000     623
## 27:  CLE 31.435897    1226
## 28:  CHN 31.851852    1720
## 29:  HOU 34.094340    1807
## 30:  CHA 35.216216    1303
## 31:  SFN 35.312500    2260
## 32:  TOR 44.785714    1254
##     team   avg_rbi tot_rbi
```

```{r}
## Same thing, but no intermediate object
## Does not order in place
res1 <- 
  baseball_dt[year > 2000, 
            list(avg_rbi=mean(rbi,na.rm=T), 
                 tot_rbi=sum(rbi,na.rm=T)), 
            by=team][order(avg_rbi)] 
```

Regression by Group
Any valid R functions can be evaluated by group….this is powerful!
We can fit a multivariate regression in the j expression and extract the p-values associated with age and mortality
The first step is to figure out how to get the result you want, then add the by= argument in to resolve the expression by group
In this case, we are fitting a linear regression, then extracting the p-values associated with the mortality predictor

```{r}
## extract the p-values for age and mortality 
cohort[,summary(lm(age ~ gender + mortality))$coefficients[2,4], 
       by=hosp_id] 
##      hosp_id          V1
##   1:      77 0.121973347
##   2:      45 0.210289025
##   3:       5 0.980924814
##   4:      13 0.887124987
##   5:      70 0.858322923
##   6:      27 0.951418487
##   7:      29 0.491992103
##   8:      85 0.239406643
##   9:      41 0.797902663
##  10:      64 0.603953309
##  11:      43 0.042534060
##  12:      62 0.288284257
##  13:      54 0.795493702
##  14:      69 0.567321025
##  15:      20 0.645297833
##  16:      46 0.282262549
##  17:      19 0.457915985
##  18:       1 0.241606281
##  19:      81 0.567469864
##  20:      55 0.630498498
##  21:      88 0.054499752
##  22:      34 0.064525949
##  23:      15 0.086299501
##  24:      35 0.410806608
##  25:      94 0.557116047
##  26:       2 0.519355657
##  27:      76 0.589302422
##  28:      75 0.206815597
##  29:      66 0.583151953
##  30:      91 0.111223202
##  31:      63 0.502365243
##  32:       7 0.550516319
##  33:      22 0.285180780
##  34:      79 0.336277925
##  35:      90 0.951291640
##  36:      95 0.541878016
##  37:      97 0.963290970
##  38:       6 0.780778184
##  39:      36 0.949266143
##  40:      61 0.250958368
##  41:      82 0.566796959
##  42:      25 0.838208685
##  43:       4 0.231794037
##  44:      50 0.333371025
##  45:      33 0.245258738
##  46:      26 0.405456544
##  47:      52 0.961953487
##  48:      12 0.975255867
##  49:      14 0.835948923
##  50:      18 0.971313166
##  51:      31 0.569980012
##  52:      53 0.646286053
##  53:      78 0.886059765
##  54:      59 0.624216847
##  55:      28 0.252300795
##  56:      93 0.020239762
##  57:      87 0.373143181
##  58:      72 0.232146768
##  59:      60 0.298261161
##  60:      51 0.794773690
##  61:      17 0.722502499
##  62:       9 0.080573683
##  63:      73 0.722582454
##  64:      39 0.330502188
##  65:       8 0.518673141
##  66:     100 0.814276270
##  67:      67 0.222394088
##  68:      56 0.011362725
##  69:      48 0.257889098
##  70:      40 0.051644566
##  71:      98 0.331692019
##  72:      57 0.860790323
##  73:      80 0.681973253
##  74:       3 0.091395810
##  75:      58 0.012353405
##  76:      74 0.260627155
##  77:      44 0.299418585
##  78:      30 0.432579802
##  79:      92 0.680032805
##  80:      96 0.032909438
##  81:      47 0.573595675
##  82:      84 0.721069120
##  83:      89 0.080750311
##  84:      24 0.118326322
##  85:      49 0.277934332
##  86:      99 0.126635305
##  87:      21 0.812356905
##  88:      83 0.351053376
##  89:      23 0.688500971
##  90:      86 0.926979107
##  91:      42 0.816799942
##  92:      16 0.809076281
##  93:      38 0.158349413
##  94:      10 0.027490918
##  95:      11 0.467260926
##  96:      32 0.241127947
##  97:      37 0.953548590
##  98:      71 0.558666806
##  99:      68 0.009686422
## 100:      65 0.055406935
##      hosp_id          V1
```

Conclusion
* Data.table functionality we covered
* Quickly read in files with fread
* Use the square brackets to interact with the data.table
* The first i argument subsets the data or joins if a data.table is given
* The second j argument accepts any valid r expressions
* If j resolves to a simple list a data.table is returned
* We can update or add columns using the := operator
* We can do any of these operations by group using the by= command
* Data.table is faster than base R in many tasks
* Data.table uses memory more efficiently by making less copies of datasets as you work with them
* Data.table is one of the fastest ways to merge and work with big data in R