---
title: "Week 01 Assignment"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

For this assignment, we will be using some simulated electronic health records (these are not real data!). This is a common sort of dataset for health care systems to use when tracking all the patients and the outpatient activity. You should take a few minutes to review the datasets using Excel, read the descriptions, and understand how they fit together. We will only use a few data files in this exercise, but you could explore all of them on your own. 
Questions
You will use these datasets to answer the questions listed below. You must be careful to think about what the appropriate denominator is for each question. As you code the answers, be mindful to use the ‘high performance’ coding approaches in data.table.
1. Are men more likely to die than women in this group of patients? Assume people without a date of death in the mortality table are still alive.
2. Are there patterns in the disease groups across gender? For every patient with at least one outpatient visit, identify if they have been diagnosed with any of the 22 conditions listed in the diseaseMap table at any time point. You will need to consider all three ICD columns in the outpatientVisit file (not just one). Create a table with the rate of disease for each condition for men, women, and all.
3. Calculate the mortality rate for every year between 2005 and 2018. Is it generally increasing, or decreasing? Assume patients are only at risk of death as of their first visit (in the outpatient Visit file). Once they have died, they are no longer at risk in subsequent year.
To submit
You can either submit the .R or .Rmd files or you can submit the link to the file in your Github repository in the text box.


```{r}
library(data.table)
library(purrr)
library(testthat)
```

```{r}
## Load the data
clinic.dt <- fread('./healthcare-data/Clinic.csv')
##Load Clinic
clinic.dt 
```

```{r}
test_that("Clinic loaded", { expect_equal(clinic.dt[,length(ClinicCode)], 60)})
test_that("Clinic loaded", { expect_equal(clinic.dt[,.N], 60)})
```

```{r}
##Load DiseaseMap
diseasemap.dt <-fread('./healthcare-data/DiseaseMap.csv')
diseasemap.dt 
```




```{r}
test_that("Disease loaded", { expect_equal(diseasemap.dt[,.N],3419)})
```

```{r}
##Load ICDCodes
icdcodes.dt <- fread('./healthcare-data/ICDCodes.csv')
icdcodes.dt 

```

```{r}
test_that("ICD10 loaded", { expect_equal(icdcodes.dt[,.N],3419)})
```

```{r}
##Load InsuranceProvider
insuranceprov.dt <- fread('./healthcare-data/InsuranceProvider.csv')
insuranceprov.dt 

```

```{r}
test_that("Insurance loaded", { expect_equal(insuranceprov.dt[,.N],15)})
```

```{r}
##Load Mortality
mortality.dt <-fread('./healthcare-data/Mortality.csv')
mortality.dt 

```

```{r}
test_that("Mortality loaded", { expect_equal(mortality.dt[,.N],7094)})
```

```{r}
##Load Outpatient Visit
outpatient.visit.dt <- fread('./healthcare-data/OutpatientVisit.csv')
outpatient.visit.dt 
```

```{r}
test_that("Outpatient loaded", { expect_equal(outpatient.visit.dt [,.N],174690)})
```

```{r}
## Load Patient
pateint.dt <- fread('./healthcare-data/Patient.csv')
pateint.dt 

```

```{r}
test_that("Patient loaded", { expect_equal(pateint.dt [,.N],20000)})
```

```{r}
#1
## Merge patient and Mortality and compare that with patient
merged_data <- merge(pateint.dt, mortality.dt, by = "PatientID", all = FALSE)  # Inner join
merged_data

```

```{r}
## results show that there is a difference - see the computations below

female.mort <- nrow(subset(merged_data, Gender == "female")) / nrow(subset(pateint.dt, Gender == "female")) *100

male.mort <- nrow(subset(merged_data, Gender == "male")) /nrow(subset(pateint.dt, Gender == "male")) *100


print(paste("female mort:", round(female.mort, 1), "%"))
print(paste("male mort:", round(male.mort, 1), "%"))

```

```{r}
menMoreLikely = female.mort < male.mort # FALSE because female mort = 35.1%; male mort = 35.9%
test_that("men more likely to die", { expect_equal(menMoreLikely, TRUE)})
```

```{r}
## 2. Combine Outputpatient with Disease on ICD10 code


# 1.Convert OutpatientVisit to long format
outpatient_long <- melt(outpatient.visit.dt, 
                        id.vars = c("VisitID", "PatientID", "VisitDate", "StaffID", "ClinicCode"), 
                        measure.vars = c("ICD10_1", "ICD10_2", "ICD10_3"), 
                        variable.name = "ICD_Type", 
                        value.name = "ICD10")

# Remove na in ICD10 
outpatient_long <- outpatient_long[!is.na(ICD10)]

# 2. Merge with DiseaseMap table to map ICD10 codes to conditions
outpatient_condition <- merge(outpatient_long, diseasemap.dt, by = "ICD10", all.x = TRUE)

# 3. Add gender data (from Patient table)
gender_data <- pateint.dt[, .(PatientID, Gender)]

# Merge gender data with outpatient condition data
outpatient_gender <- merge(outpatient_condition, gender_data, by = "PatientID", all.x = TRUE)

# 4. Calculate disease rates by condition and gender
disease_rates <- outpatient_gender[, .(
  Total = uniqueN(PatientID),
  Male = uniqueN(PatientID[Gender == "male"]),
  Female = uniqueN(PatientID[Gender == "emale"])
), by = Condition]

# Calculate the percentage rates for males, females, and overall
disease_rates[, `:=`(Rate_Male = round(100 * Male / Total, 2),
                     Rate_Female = round(100 * Female / Total, 2),
                     Rate_All = round(100 * Total / sum(Total), 2))]

# 5. print the result 
summaries.dt <- disease_rates[, .(Condition, Rate_Male, Rate_Female, Rate_All)]
print(summaries.dt)


# Remove rows where Condition is NA
summaries.dt <- summaries.dt[!is.na(Condition)]

# Print the final result
print(summaries.dt)
```

```{r}
summaries.dt

```

```{r}
test_that("summaries populated", { expect_equal(summaries.dt [,.N],22)})
```

```{r}

## 3.
str(mortality.dt)
mortality.dt

str(pateint.dt)
```




```{r}

#get_interval <- function(yr){
#  df 
#  return(df)
#}


# yr <- 2013
# get_interval(yr)
#
# deaths.dt <- mortality.dt[DateOfDeath >= eval(get_interval(yr)$start.date) &
#                             DateOfDeath <= eval(get_interval(yr)$end.date)]
#
# patient.dt[deaths.dt, on="PatientID"]
#
# nrow(patient.dt[deaths.dt, on="PatientID"])/nrow(patient.dt) * 100
#
# ## compute the alive records - all those whose death was during or after the current year
# alive.dt <- mortality.dt[DateOfDeath >= eval(get_interval(yr)$start.date)]
#
# nrow(alive.dt[deaths.dt, on="PatientID"])/nrow(alive.dt) * 100

#get_mortality_rate <- function(yr){

#  mortality_rate 
#  return(mortality_rate)
#}

#get_mortality_rate(2013)


```

get_mortality_rate function:

1.計算當年死亡人數：篩選死亡日期在該年範圍內的記錄
計算風險人群：

2.找出每個病人的第一次就診日期
篩選第一次就診日期在目標年份或之前的病人
排除該年之前已死亡的病人

3.計算死亡率：當年死亡人數/風險人群數量 * 100

關鍵點是風險人群的定義：必須是在該年之前有就診記錄且年初還活著的病人。




```{r}


get_interval <- function(yr) {
  start.date <- as.Date(paste0(yr, "-01-01"))
  end.date <- as.Date(paste0(yr, "-12-31"))
  list(start.date = start.date, end.date = end.date)
}


get_mortality_rate <- function(yr) {
 interval <- get_interval(yr)
  
  # Get deaths in current year
  deaths.dt <- mortality.dt[DateOfDeath >= interval$start.date & 
                          DateOfDeath <= interval$end.date]
  
  # Get patients who had their first visit before or during this year
  at_risk.dt <- outpatient.visit.dt[, .(first_visit = min(VisitDate)), by = PatientID][
    first_visit <= interval$end.date
  ]
  
  # Get patients who died before this year (no longer at risk)
  prior_deaths.dt <- mortality.dt[DateOfDeath < interval$start.date]
  
  # Remove prior deaths from at-risk population
  current_at_risk <- at_risk.dt[!prior_deaths.dt, on = "PatientID"]
  
  # Calculate mortality rate
  mortality_rate <- nrow(deaths.dt) / nrow(current_at_risk) * 100
  
  return(mortality_rate)
}

```


```{r}

years <- 2005:2018
mortality_rates <- sapply(years, get_mortality_rate)
mortality_rate_table <- data.table(Year = years, MortalityRate = mortality_rates)
print(mortality_rate_table)

```





```{r}
test_that("2013 is correct", { expect_equal(get_mortality_rate(2013), 1.638896)})
```



